Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-09-04 docs/other_docs/manuscript_comments/parasite_mortality_manuscript_current.tex
Rearranged format of directories

\abstract{

Parasites can often have significant impacts on host populations through parasite-induced host mortality.  Detecting and quantifying the magnitude of parasite-
induced mortality, particularly for macroparasites where pathology is
linked to to the intensity of infection, can provide important insight into
parasite transmission and host population management. This is, however,
notoriously difficult in practice as the only data often available are measures of parasite intensities. In this study we provide a novel method for detecting and
quantifying parasite-induced mortality in wildlife populations from intensity
data and, using simulations, we show that our method is more reliable for
detecting and quantifying parasite-induced host mortality than the methods that are
currently available.  We also show that this method provides quantitative
assessments of parasite-induced mortality in empirical data that are consist
with previously published qualitative assessments. However, we stress that this
method, along with all methods for estimating parasite-induced mortality in
host populations, have a number critical assumptions that limit their
applicability in the real world.  We conclude that methods for estimating and
detecting parasite-induced mortality from intensity data should only be used as an
exploratory tool for informing more rigorous studies of parasite-induced host
mortality.}

@@

Infectious agents can have major impacts on animal populations through changing
population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies (Coyne et al. 1989) [get citation], bovine TB \citep{Cox2005}, and
rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
disease, and does not account for the number of infectious agents present.
This method is sufficient for many bacterial and viral agents that reproduce within a
host, however for macroparasites pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013},
wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
wild host populations must succumb to parasitic infections, it is notoriously
difficult to actually quantify parasite-induced host mortality (PIHM) in wild
animal populations \citep{McCallum2000a}.

Ideally, parasite-induced host mortality is
quantified by experimentally infecting and tracking individual hosts in the
wild population; however, for logistical and ethical reasons this method is
rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
relationship between infection intensity and host mortality \citep{Adjei1986}.

\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
observed parasite distribution in the host population to the distribution
predicted in the absence of parasite-induced mortality. We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1. This method
assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts. Assuming mortality occurs only in these heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
fitting a negative binomial distribution to hosts with lower and lower parasite
loads, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data.

The Crofton Method may be able to detect the presence of PIHM, but it does not quantify the relationship between infection intensity and host survival
probability. \cite{Adjei1986} suggested that this relationship
could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using this distribution to calculate the
probability of host survival with increasing parasite intensity. To do this,
\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
This method appeared to provide an estimate for the parasite intensity at which
50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
population that was lost (\emph{SI} 2). However, to implement this method the observed data must be
modified to fit the GLM framework and subjectively binned when mean
infection intensity is high or sample sizes are small.

After 30 years, and despite clear limitations \citep{McCallum2000a}, these
methods (particularly the Crofton Method) are still discussed among
parasitologists and are the primary techniques for examining population level
impacts of parasitism using parasite intensity data. In these methods, PIHM can
only be identified by visually examining plots and, with no clear decision
rule, it can be difficult to determine the significance of PIHM across
different host-parasite systems. The survival function given by the Adjei
Method offers one solution; however, this method requires manipulating the
original data and has never been tested.

Intensity data should only
be used to estimate parasite impacts on host populations if unbiased and accurate methods exist. In
this study, we first propose a novel method for detecting and quantifying PIHM. We
next use simulations to compare our method with the previous Adjei
Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the
lethal parasite load ($LD_{50}$) and the associated survival function.  We then
apply both methods to real datasets previously used in PIHM analyses and
compare the results. Finally, we discuss the limitations of inferring PIHM
from intensity data and whether any method for inferring PIHM has a place
in quantitative parasitology.

@@

Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
require binning or data alteration, reduces the number of
parameters to be estimated, and  uses standard statistical techniques to determine PIHM significance. We provide Python code for implementing the Likelihood Method in \emph{SI} 4.

As with all previously proposed methods for estimating PIHM, the
Likelihood Method first assumes that prior to mortality the parasite distribution is described by a negative binomial $g(x; \mu_p, k_p)$, where $\mu_p$ and $k_p$ are the mean parasite intensity and aggregation before mortality, respectively (smaller $k_p$ indicates more aggregation).   Previous methods required calculating the total number of hosts before mortality ($N_p$) \citep{Crofton1971a,Adjei1986}, however this parameter is not needed in the Likelihood Method.

The Likelihood Method then assumes that the host survival function, which specifies the probability of a host surviving with $x$ parasites, follows the logistic curve given by

\begin{equation}
    h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$ by  first defining a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.  Using standard rules of conditional probability this distribution can be written as

@@

One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$. Therefore equation \ref{eq:concept} can be written as

\begin{equation}
    P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
    \label{eq:dist}
\end{equation}

Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset. The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.  All parameters are defined in Table \ref{tab:params}.

To estimate the significance of PIHM in a host-parasite system, a
likelihood ratio test is used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by a negative binomial
distribution.  If PIHM is not significant in the system, the resulting likelihood
ratio statistic should approximately follow a $\chi^2$ distribution with two degrees of freedom.

@@

We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters. First, we created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$. Second, we chose values of $a$ and $b$ and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
between 0 and 1 and if the calculated host survival probability was less than this random
number, the host experienced parasite-induced mortality. The parasite distribution in these simulated surviving hosts is equivalent to the observed parasite distribution in a wild host population that has undergone parasite-induced host mortality.

@@

We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had graduate, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).  For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500]. $N_p$ is not technically the sample size on which the methods are being
tested on the post-mortality data because PIHM reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.\\

\noindent
\emph{Question 2: Can we estimate fatal parasite intensity and the host survival function?}

To compare the ability of the Adjei Method and the Likelihood Method to recover the $LD_{50}$ and the parameters $a$ and $b$ of the survival function, we used the same simulation procedure and parameter combinations described above. For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates over pre-mortality host population sizes of $N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]. We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.  Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.

\subsection*{Efficacy of the Likelihood Method with unknown pre-mortality parameters}

In the final simulation, we tested the ability of the Likelihood Method to
correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality
parameters are unknown. The previous simulations showed that the Likelihood
Method effectively identified PIHM when $\mu_p$ and $k_p$ were known with
values of 10 and 1, respectively.  As a best-case scenario, we simulated host-
parasite systems with these pre-mortality parameters and tested the power of
the Likelihood Method to identify PIHM for gradual, moderate and steep survival
functions when the pre-mortality parameters $\mu_p$ and $k_p$ also needed to be estimated.  We
perform 500 simulations over a range of different samples sizes following the
simulation procedure described above.

@@

In both earlier studies, the authors reported PIHM in some of the datasets and we tested whether the Adjei
Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
Method to estimate the pre-mortality distribution, and then ran the Likelihood
Method and Adjei Method using these pre-mortality parameters.  For the
\cite{Adjei1986} datasets, we followed the same procedure as the authors and
first truncated the data at 2 parasites and then fit the Crofton Method for the
female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
distributions for each species with the results from the females.  Finally, we
applied the Adjei Method and the Likelihood Method to determine whether or not
PIHM was significant for these species and compared our results to those given by the authors.  All fitting to data was done with the code provided in \emph{SI} 4.

>>>>

\section*{Abstract}

Parasites can often have significant impacts on host populations through parasite-induced host mortality.  Detecting and quantifying the magnitude of parasite-
induced mortality, particularly for macroparasites in which pathology is
linked to to the intensity of infection, can provide important insight into
parasite dynamics and host population management. This is, however,
notoriously difficult in practice as the only data often available are measures of parasite intensities. In this study we provide a novel method for detecting and
quantifying parasite-induced mortality in wildlife populations from intensity
data and, using simulations, we show that our method is more reliable for
detecting and quantifying parasite-induced host mortality than the methods that are
currently available.  We also show that this method provides quantitative
estimates of parasite-induced mortality in empirical data that are consist
with previously published qualitative estimates. However, we stress that this
method, along with all methods for estimating parasite-induced mortality in
host populations for intensity data alone, has a number critical assumptions that limit their
applicability in the real world.  We conclude that methods for estimating and
detecting parasite-induced mortality from intensity data should be used only as an
exploratory tool for informing more rigorous studies of parasite-induced host
mortality.

@@

Infectious agents can have major impacts on animal populations through changing
population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of microparasite pathogens, such as rabies \citep{Coyne1989}, bovine TB \citep{Cox2005}, and
rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
disease, and does not account for the number of infectious agents present.
This method is sufficient for many bacterial and viral agents that reproduce within a
host, however for macroparasites, pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and
wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
wild host populations must succumb to parasitic infections, it is notoriously
difficult to actually quantify parasite-induced host mortality (PIHM) in wild
animal populations \citep{McCallum2000a}.

Ideally, parasite-induced host mortality is
quantified by experimentally infecting and tracking individual hosts in the
wild population; however, for logistical and ethical reasons this method is
rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and to quantify the
relationship between infection intensity and host mortality \citep{Adjei1986}.

\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
observed parasite distribution in the host population to the distribution
predicted in the absence of parasite-induced mortality. We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1. This method
assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts. Assuming mortality occurs only in heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
fitting a negative binomial distribution to hosts with lower and lower parasite
loads, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data. [FIGURE]

The Crofton Method may be able to detect the presence of PIHM, but it does not quantify the relationship between infection intensity and host survival
probability. \cite{Adjei1986} suggested that this relationship
could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using this distribution to calculate the
probability of host survival with increasing parasite intensity. To do this,
\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
\citeauthor{Adjei1986} suggested that this method could provide an estimate for the parasite intensity at which a host has a 50\% chance of suffering parasite-induced mortalith ($LD_{50}$). However, to implement this method the observed data must be
modified to fit the GLM framework and subjectively binned when mean
infection intensity is high or sample sizes are small (see SI 2 for details).

After 30 years, and despite clear limitations \citep{McCallum2000a}, these
methods (particularly the Crofton Method) are still discussed among
parasitologists and are the primary techniques for examining population level
impacts of parasitism using parasite intensity data. In these methods, PIHM can
only be identified by visually examining plots and, with no clear decision
rule, it can be difficult to determine the significance of PIHM across
different host-parasite systems. The survival function given by the Adjei
Method offers one solution; however, this method requires manipulating the
original data and its accuracy has never been validated.

Intensity data should
be used to estimate parasite impacts on host populations only if unbiased and accurate methods exist. In
this study, we first propose a novel method for detecting and quantifying PIHM. We
next use simulations to compare our method with the previous Adjei
Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the
lethal parasite load ($LD_{50}$) and the associated survival function.  We then
apply both methods to real datasets previously used in PIHM analyses and
compare the results. Finally, we discuss the limitations of inferring PIHM
from intensity data and whether any method for inferring PIHM has a place
in quantitative parasitology.

@@

Here we propose a novel, likelihood-based method (henceforth Likelihood Method)
that does not require binning or data alteration, reduces the number of
parameters to be estimated, is highly generalizable, and uses standard
statistical techniques to determine PIHM significance.  The Likelihood Method
begins with the same assumptions as the Adjei Method: namely that infection of
a host, parasite-induced mortality of a host, and the sampling of a host
population occur at distinct time intervals during a host's life. As discussed by
\citeauthor{Adjei1986}, this is not necessarily unrealistic as studies have
shown that infection and host mortality are often age and/or body size dependent [citations].

The Likelihood Method then assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \boldsymbol{\phi})$, which specifies the probability of a host having $x$ parasites when it is observed.  $\boldsymbol{\phi}$ is a vector of parameters that describes the shape of this distribution. The method then assumes that the probability of a host surviving with $x$ parasites from infection until sampling is given by $h(\text{survival} ; x, \boldsymbol{\theta})$ where $\boldsymbol{\theta}$ specifies any additional parameters needed to define the host survival function.

With these two assumptions, we can define a distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival, $P(x | \text{survival})$.  Using standard rules of conditional probability this distribution can be written as

@@

$P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \boldsymbol{\theta})$, $P(x)$ is the pre-mortality parasite distribution $g(x; \boldsymbol{\phi})$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})$. Therefore equation \ref{eq:concept} can be written as

\begin{equation}
    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}{\sum_{x=0}^{\infty} h(\text{survival}; x, \boldsymbol{\theta})  * g(x; \boldsymbol{\phi})}
    \label{eq:dist}
\end{equation}

% \begin{equation}
%     h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
%     \label{eq:logistic}
% \end{equation}

Using this probability distribution, one can then find the parameters $\boldsymbol{\theta}$
and $\boldsymbol{\phi}$ that maximize the likelihood of an observed host-parasite dataset.
To estimate the significance of PIHM in a host-parasite system, a likelihood
ratio test can be used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
$g(x; \boldsymbol{\phi})$.  If PIHM is not significant in the system, the resulting
likelihood ratio statistic should approximately follow a $\chi^2$ distribution
with degrees of freedom equal to the number of parameters in the full model with parasite-induced mortality minus the number of parameters in the reduced model without parasite-induced mortality [citation].

Equation \ref{eq:dist} could be parameterized in many different ways depending
on the parasite system of interest. In this study, we adopt the typical assumption that the pre-
mortality parasite distribution $g(x; \boldsymbol{\phi})$ follows a negative binomial
distribution with the parameters mean parasite intensity ($\mu_p$) and
aggregation ($k_p$) before mortality (smaller $k_p$ indicates
more aggregation) \citep{Crofton1971a,AndersonandMay1978,Adjei1986}.   The negative binomial distribution can arise as the equilibrium parasite
distribution under a variety of different biological and statistical assumptions
\citep{Kendall1948a, Boswell1970, Calabrese2011}. However, it is also an incredibly flexible
distribution that fits many host-parasite systems regardless of whether the
underlying mechanisms lead to an exact negative binomial distribution
\citep{Shaw1998}.

Choosing an appropriate function for $h(\text{survival}; x, \boldsymbol{\theta})$ will depend on the system under consideration.  Many theoretical models of
parasite-induced host mortality assume that the parasite-induced death rate of
hosts is a linear function of parasite intensity
\citep{AndersonandMay1978,Dobson1992,Barbour2000}. It has been previously noted that parasite-induced mortality can be nearly impossible to detect from intensity data when the host survival function is a linear function of parasite intensity as the post-mortality distribution will be of a similar form as the pre-mortality distribution \citep{Lanciani1989}.  That being said there is substantial empirical evidence for non-linear host-survival functions  \citep{Benesh2011}.

As one of the goals of this study is to compare this new
Likelihood Method to the previously proposed Adjei Method, we adopt the host-
survival function used in their study and assume host-survival is non-linear and follows a logistic function given by

\begin{equation}
    h(\text{survival}; x, a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
where $b / 4$ determines the maximum rate of decline of
host survival probability with increasing parasite load, analogous to the pathogenicity parameter $\alpha$ in traditional macroparasite models \citep{AndersonandMay1978}.  When $b$ is held constant, for every one unit increase in $a$ the parasite intensity at which 99\% of hosts survive increases by $ 1/ b$.  The equation $\exp(a
/ b)$ can also be used to calculate the parasite $LD_{50}$, here defined as the
infection intensity at a host has a 50\% probability of dying.  This function is commonly used in toxicology and survival analysis and has the useful properties of being bounded between 0 and 1 and being differentiable for all $x$.  That being said, it is phenomenological and there is little theoretical justification to use it rather than it tends to fit survival data. However, given that a goal of these analyses is to
compare this method's results to the Adjei Method, to facilitate comparison it is natural
to adopt the same host-survival function.  When
applying the likelihood method to other systems, other, more mechanistic host-survival functions can be used in place of equation \ref{eq:logistic}.

@@

We tested the ability of the Adjei and the Likelihood Methods to identify the presence of PIHM on simulated data with known pre-mortality parameters. Consistent with the assumptions of the model that parasite infection, mortality, and sampling occur at distinct life stages of the host, we first created a pre-mortality host population by drawing $N_p$ randomly infected hosts from a
negative binomial distribution with parameters $\mu_p$ and $k_p$. This is equivalent to the period of hosts becoming infected with parasites given in \cite{Adjei1986}.  In the Adjei Method and Crofton Method, $N_p$ is a necessary parameter that is defined as the number of hosts in the population before parasite-induced mortality.  A more appropriate way to define this parameter is the number of hosts that would have been sampled had parasite-induced host mortality not occurred.  This parameter is not necessary when using the Likelihood Method because unlike the Adjei Method and Crofton Method which estimate parasite-induced mortality using absolute numbers of hosts, the Likelihood Method estimates parasite-induced mortality using probabilities. However, to compare the results of the Likelihood Method with the Adjei Method, we specified a value for $N_p$ for all simulations.

Second, we chose values of $a$ and $b$ for the host survival function and calculated the probability of survival
for all $N_p$ hosts using equation \ref{eq:logistic}.  Then, for each host, we drew a random number from a uniform distribution
between 0 and 1 and if the calculated host survival probability was less than this random
number, the host experienced parasite-induced mortality.  This was the period in which hosts died due to infection. The parasite distribution in these simulated surviving hosts represented the parasite distribution in a wild host population that has undergone parasite-induced host mortality.

@@

We used three different values of $\mu_p$ (10, 50, 100) and for each $\mu_p$ we examined three different survival functions that had gradual, moderate, and steep decreases in the host survival with increasing parasite intensity (Figure \ref{fig:question1}A).  For a given $\mu_p$, each survival function had the same $LD_{50}$ ([$\mu_p = 10$, $LD_{50} = 7.39$], [$\mu_p = 50$, $LD_{50} = 35.57$], [$\mu_p = 100$, $LD_{50}= 77.3$]),  but different values of $a$ and $b$.  We examined each $\mu_p$-survival function pair at  three levels of parasite
aggregation, $k_p = 0.1$, 0.5, and 1 --- realistic values of parasite aggregation in natural populations \citep{Shaw1998}.  For each of these parameter
combinations we simulated 150 datasets and tested the probability of each method correctly identifying PIHM in the post-mortality dataset (power) and incorrectly identifying PIHM in the pre-mortality dataset (Type I error).  For each method, we used a likelihood ratio test to determine whether the full model with PIHM provided a significantly better fit than the reduced model without PIHM at significance level of 0.05.  We tested each parameter combinations for pre-mortality population sizes of $N_p$ = [50, 100, 200, 300, 400, 500]. $N_p$ is not technically the sample size on which the methods are being
tested for the post-mortality data because PIHM reduces $N_p$ for each simulated
dataset.  We therefore used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size in the power simulations.

In the next simulation, we tested the ability of only the Likelihood Method to
correctly identify PIHM and estimate $LD_{50}$ when the pre-mortality
parameters are unknown.  As a best-case scenario, we simulated host-
parasite systems with $\mu_p = 10$ and $k = 1, 0.5, \text{and} 0.1$ [re-run for 0.5 and 0.1] as it is easier to detect PIHM from small samples sizes when mean parasite intensity is low. We then used
the Likelihood Method to identify PIHM for gradual, moderate and steep survival
functions when the pre-mortality parameters $\mu_p$ and $k_p$ also needed to be estimated.  We
perform 500 simulations over a range of different samples sizes following the
simulation procedure described above. \\

\noindent
\emph{Question 2: Can we estimate properties of the host survival function?}

In the previous section we compared the ability of the Adjei Method and the Likelihood Method to correctly identify whether or not PIHM was a occurring in a system (i.e. a yes or no answer). In this section we compare the ability of the Adjei Method and the Likelihood Method to estimate properties of the survival function such as the parameters $a$, $b$ and $LD_{50}$.  To do this, we used the same simulation procedure and parameter combinations described above. For each parameter
combination we simulated 150 datasets, estimated $a$, $b$, and $LD_{50}$ and calculated the standardized bias and
precision \citep{Walther2005} for these estimates.  Because estimating properties of the host survival function requires more information than simply detecting PIHM, we used larger values of $N_p$ for this simulation ($N_p$ = [300, 500, 1000, 2000, 5000, 7500,
10000]).  We used the average number of surviving hosts over all 150 simulations for a given parameter combination as our measure of sample size.  Because parameters $a$ and $b$ showed similar patterns of bias and precision, we only show the results for $a$.

@@

In both earlier studies, the authors reported PIHM in some of the datasets and we tested whether the Adjei
Method and/or the Likelihood Method also predicted PIHM. For the 6 datasets from
\cite{Crofton1971a}, we truncated the data at 4 parasites, applied the Crofton
Method to estimate the pre-mortality distribution, and then ran the Likelihood
Method and Adjei Method using these pre-mortality parameters.  For the
\cite{Adjei1986} datasets, we followed the same procedure as the authors and
first truncated the data at 2 parasites and then fit the Crofton Method for the
female fish of both species.  Then, following the original authors' methods, we parameterized the male pre-mortality
distributions for each species with the results from the females.  Finally, we
applied the Adjei Method and the Likelihood Method to determine whether or not
PIHM was significant for these species and compared our results to those given by the authors.  All code for the analyses is provided in \emph{SI} 4.
